---
title: 'Class 17: Vaccination Rate Mini Project'
author: "Tasnia Sharia (PID A15931128)"
date: "11/23/2021"
output:
  html_document: default
  pdf_document: default
---

# Background

The goal of this hands-on mini-project is to examine and compare the Covid-19 vaccination rates around San Diego.

We will take data from the CA state gov website to retrieve our data:  https://data.ca.gov/dataset/covid-19-vaccine-progress-dashboard-data-by-zip-code

# Getting Started

```{r}
# Import vaccination data
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```

> **Q1. What column details the total number of people fully vaccinated?**

```{r}
#vax$persons_fully_vaccinated

#hiding the output so it does not take up s many pages when submitting
```

"persons_fully_vaccinated" column details the total number of people fully vaccinated

> **Q2. What column details the Zip code tabulation area?**

```{r}
#vax$zip_code_tabulation_area

#hiding the output so it does not take up s many pages when submitting
```

"zip_code_tabulation_area" details the Zip code tabulation area.

> **Q3. What is the earliest date in this dataset?**

```{r}
head(vax$as_of_date)
```
 
 January 5, 2021 is the earliest data
 
 > **Q4. What is the latest date in this dataset?**
 
```{r}
tail(vax$as_of_date)
```
 
November 16, 2021 is the latest date

```{r}
#getting quick overview of the dataset
#had to install skimr package
skimr::skim(vax)
```

> **Q5. How many numeric columns are in this dataset?**

There are 9 numeric columns.

> **Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column?**

There are 8256 NA values in the persons_fully_vaccinated column.

> **Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)?**

```{r}
8256/81144 * 100
```
10.17% of persons fully vaccinated values are missing.

> ** [Optional]: Why might this data be missing?**

Many people chose to not get vaccinated or felt that the first dose is effective enough.

 **Working with Dates**
 
 The lubridate packages can make working with dates and times a lot easier.
 
```{r}
#had to install the lubridate package
#loading the library
library(lubridate)
```

```{r}
#what is the date today
today()
```
```{r}
# Specify that we are using the Year-month-day format easier for the lubridate package
vax$as_of_date <- ymd(vax$as_of_date)
```

```{r}
today() - vax$as_of_date[1]
```
```{r}
#We can now determine how many days the dataset span
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```
 
 > **Q9. How many days have passed since the last update of the dataset?**
 
```{r}
today() - vax$as_of_date[81144]
```
 
 > **Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)?**
 
```{r}
length(unique(vax$as_of_date))
```

# Working with Zip Codes
 
 We can use the zipcodeR package to make working with zip code column easier
 
```{r}
#had to install zipcodeR package
#loading library
library(zipcodeR)
```
 
```{r}
geocode_zip('92037')
```
 
```{r}
#Calculating the distance between the center of any two ZIP codes in miles
zip_distance('92037','92109')
```
 
```{r}
#can pull census data about ZIP code areas (like median household income)
reverse_zipcode(c('92037', "92109") )
```

```{r}
#pulling data for all ZIP codes in the dataset
zipdata <- reverse_zipcode( vax$zip_code_tabulation_area )
```

# Focus on the San Diego area

We will focus on the San Diego County area by referring to vax$county == "San Diego" entries.

```{r}
# Subset to San Diego county only areas using R code
#sd <- vax[vax$county == "San Diego", ]
```

```{r}
#or we can use the dplyr package which is more convenient when we are subsetting across multiple criteria 
library(dplyr)
sd <- filter(vax, county == "San Diego")
nrow(sd)
```

> **Q11. How many distinct zip codes are listed for San Diego County?**

```{r}
length(unique(sd$zip_code_tabulation_area))
```

> **Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset?**

```{r}
ind <- which.max(sd$age12_plus_population)
sd[ind,]

```

> **Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2021-11-09”?**

```{r}
sd.row <- filter(sd, as_of_date=="2021-11-09")
mean(sd.row$percent_of_population_fully_vaccinated, na.rm=TRUE)
```

The overall average is 67.3%

> **Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2021-11-09”?**

```{r}
hist(sd.row$percent_of_population_fully_vaccinated)
```

```{r}
#making a plot using ggplot
#loading library
library(ggplot2)

ggplot(sd.row) + aes(percent_of_population_fully_vaccinated) + geom_histogram(bins=15)
```

# Focus on UCSD/La Jolla

```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```

> **Q15. Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area:**

```{r}
filter(sd.row, zip_code_tabulation_area=="92037")
```


```{r}
ggplot(ucsd) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated")
```

Population in the 92037 ZIP code area
```{r}
ucsd[1,]$age5_plus_population
```

# Comparing 92037 to other similar sized areas?

Looking across every zip code area with a population at least as large as that of 92037 on as_of_date “2021-11-16”

```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2021-11-16")

#head(vax.36)
```


> **Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2021-11-16”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?**

```{r}
ggplot(ucsd) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  geom_hline(yintercept = 0.67276, color="red") +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated")
```

> **Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2021-11-16”?**

```{r}
summary(vax.36$percent_of_population_fully_vaccinated)
```

> **Q18. Using ggplot generate a histogram of this data.**

```{r}
ggplot(vax.36) + aes(percent_of_population_fully_vaccinated) + geom_histogram() + labs(x="Percent Vaccinated", y="Count")
```

> **Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?**

```{r}
vax %>% filter(as_of_date == "2021-11-16") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)
```

```{r}
vax %>% filter(as_of_date == "2021-11-16") %>%  
  filter(zip_code_tabulation_area=="92109") %>%
  select(percent_of_population_fully_vaccinated)
```

The zip code 92040 is below the average and the zip code 92109 is above average.

> **Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144.**

```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)
```

```{r}
ggplot(vax.36.all) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color="blue") +
  labs(x="Date", y="Percent Vaccinated",
       title="Vaccination Rate Across California",
       subtitle="Only areas with a population above 36% are shown") +
  geom_hline(yintercept = 0.66, col="red")
```

> **Q21. How do you feel about traveling for Thanksgiving and meeting for in-person class next Week?**

Many students would travel during the break.  It would be safer is classes were remote afterwards.