---
title: 'Class 6: R Functions'
author: "Tasnia Sharia (PID A15931128)"
date: "10/14/2021"
output:
  pdf_document: default
  html_document: default
---

## A play with Rmarkdown
This is some plain text. I can make things **bold**. I can also make *things italic*.

```{r}
# This is a code chunk
plot(1:10)
```

##R functions

In today's class we are going to write a function together that grades some students work.

Questions for today's:

> **Q1.** Write a function grade() to determine an overall grade from a vector of student homework assignment scores dropping the lowest single score. If a student misses a homework (i.e. has an NA value) this can be used as a score to be potentially dropped. Your final function should be adquately explained with code comments and be able to work on an example class gradebook such as this one in CSV format: “https://tinyurl.com/gradeinput” [3pts]

```{r}
# Example input vectors to start with
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
```

Let's start with student1 and find their average score.
```{r}
mean(student1)
```
But we want to drop the lowest score...
We could try the **min()** function
```{r}
min(student1)
```
The **which.min** function looks useful:
```{r}
which.min(student1)
```
This gives the position of the lowest score.
```{r}
# This would be the lowest score for student1
student1[ which.min(student1) ]
```
To drop this value I can use minus
```{r}
student1[ -which.min(student1) ]
```
Now we can find the mean() with the lowest score dropped
```{r}
mean(student1[ -which.min(student1) ])
```

Now let's try student2 grades
```{r}
student2
```
This did not work cause of the NA
```{r}
mean(student2[ -which.min(student2) ])
```
We need to remove the NA elements of the student2 vector
```{r}
#which.min(student2)
mean(student2[ -which.min(student2)], na.rm=TRUE )
```
It dropped the 80(the lowest score) but not the NA(missing homework).  Need to try something else

Let's try student3
```{r}
student3
```
```{r}
mean(student3[ -which.min(student3)], na.rm=TRUE )
```

One new idea/approach is we could replace the NA (missing homework) with zero

Let's go back to student2
```{r}
student2
```

```{r}
is.na(student2)
```
The **is.na()** functions returns a logical vector where TRUE elements represent where the NA values are positioned.

```{r}
which(is.na(student2))
```

Now let's make the NA values into zeros
```{r}
student.prime <- student2
student.prime
student.prime[ which(is.na(student.prime)) ] = 0
student.prime
```

Now we need to put this all together to get the average score dropping the lowest score where they map the NA values to zero.
```{r}
student.prime <- student2
student.prime[ which(is.na(student.prime)) ] = 0
mean(student.prime[ -which.min(student.prime) ])
```
Let's check if it works for student3
```{r}
student.prime <- student3
student.prime[ which(is.na(student.prime)) ] = 0
mean(student.prime[ -which.min(student.prime) ])
```
We got our working snippet! Let's simplify.
```{r}
x <- student3
# Map NA values to zero
x[ which(is.na(x)) ] = 0
# FInd the mean without the lowest value
mean(x[ -which.min(x) ])
```

Now we can use this as the body of my functions and create grade()
```{r}
grade <- function(x) {
  # Make sure our scores are all numbers
  x <- as.numeric(x)
  # Map NA values to zero
  x[ which(is.na(x)) ] = 0
  # Find the mean without the lowest value
  mean(x[ -which.min(x) ])
}
```

```{r}
grade(student2)
```

Now read the full grade book CSV file.
```{r}
scores <- read.csv("https://tinyurl.com/gradeinput", row.names=1)
scores
```

Making the NA inputs into numeric values.
Had to change the grade() by adding the as.numeric() 
```{r}
grade(as.numeric(scores[10,]))
```

```{r}
is.numeric(scores[10,])
```

Now grade all students by using the **apply()** function.
```{r}
ans <- apply(scores,1,grade)
ans
```

> **Q2.** Using your grade() function and the supplied gradebook, Who is the top scoring student
overall in the gradebook? [3pts]

```{r}
which.max(ans)
```


> **Q3.** From your analysis of the gradebook, which homework was toughest on students (i.e. obtained the lowest scores overall? [2pts]

We can use the apply() function over the columns by setting the margin=2 argument.
```{r}
apply(scores,2,mean, na.rm=TRUE)
lowHW <- which.min(apply(scores,2,mean, na.rm=TRUE))
lowHW
```

> **Q5.** Make sure you save your Rmarkdown document and can click the “Knit” button to generate a PDF foramt report without errors. Finally, submit your PDF to gradescope. [1pt]


> **Q4.** Optional Extension: From your analysis of the gradebook, which homework was most predictive of overall score (i.e. highest correlation with average grade score)? [1pt]

```{r}
y <- scores
# maps NA values to zero
y[is.na(y)] = 0
y

cor(ans, scores$hw1)
```
Applying correlation to each hw
```{r}
apply(y,2,cor,ans)
```

Finding the highest correlation
```{r}
which.max(apply(y,2,cor,ans))
```

We want to optionally turn of dropping the lowest score for grading
```{r}
grade2 <- function(x, drop.lowest=TRUE) {
  # Make sure our scores are all numbers
  x <- as.numeric(x)
  # Map NA values to zero
  x[ which(is.na(x)) ] = 0
  
  
  if(drop.lowest){
    # Find the mean without the lowest value
    mean(x[ -which.min(x) ] )
  } else {
    mean(x)
  }
  
}
```


